Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-23 paper-new/appendix.tex
sketchy stuff in the appendix on syntax and update to the extensional model

>>>>

\section{Gradual Typing Syntax}

\max{decide if we want to include $\times$ or not}

\begin{align*} % TODO is hole a term?
  &\text{Types } A := \dyn \alt \nat \alt A \times A \alt (A \ra A') \\
  &\text{Value Contexts } \Gamma := \cdot \alt (\Gamma, x : A) \\
  &\text{Terms } M, N := \upc c M \alt \dnc c M
  &\quad\quad \alt \ret {V} \alt \bind{x}{M}{N} \alt V_f\, V_x \alt \dn{A}{B} M 
\end{align*}

The type precision derivations $c : A \ltdyn A'$ is inductively defined by
reflexivity, transitivity, congruence for $\ra$ and $\times$, and generators
$\textsf{Inj}_\ra : (D \ra D) \ltdyn D$ and $\textsf{Inj}_{\text{nat}}
: \nat \ltdyn D$.
%
We define equivalence of type precision derivations to be inductively generated by congruence for all constructors, category laws for reflexivity and transitivity as well as functoriality laws for $\ra$ and $\times$ congruence
\[ (c_i \ra c_o)(c_i' \ra c_o) = (c_ic_i') \ra (c_oc_o') \]
\[ (c_1 \times c_2)(c_1' \times c_2') = (c_1c_1') \times (c_2c_2') \]

\begin{theorem}
  \begin{enumerate}
  \item For every $A$, there is a derivation $c : A \ltdyn D$
  \item Any two derivations $c,c' : A \ltdyn A'$ of the same precision
    are equivalent.
  \end{enumerate}
\end{theorem}
\begin{proof}
  \begin{enumerate}
  \item See prior work \cite{prior-work}
  \item We show this by showing that derivations have a canonical
    form.

    The following presentation of precision derivations has unique derivations
    \begin{mathpar}
      \inferrule{}{\textrm{refl}(D) : D \ltdyn D}
      \inferrule{}{\textsf{Inj}_{\text{nat}} : \nat \ltdyn D}
      \inferrule{}{\textrm{refl}(\nat) : \nat \ltdyn \nat}
      \inferrule{c : A_i \ra A_o \ltdyn D\ra D}{c(\textsf{Inj}_{\text{arr}}) : A_i \ra A_o \ltdyn \nat}
      \inferrule{c : A_i \ltdyn A_i' \and d : A_o \ltdyn A_o'}{c \ra d : A_i \ra A_o \ltdyn A_i'\ra A_o'}
      %% \inferrule{A_1 \times A_2 \ltdyn D\times D}{A_1 \times A_2 \ltdyn \nat}
      %% \inferrule{A_1 \ltdyn A_1' \and A_2 \ltdyn A_2'}{A_1 \times A_2 \ltdyn A_1'\times A_2'}
    \end{mathpar}
    Since it satisfies reflexivity, cut-elimination and congruence, it
    is a model of the original theory. Since it is a sub-theory of the
    original theory, it is equivalent.
  \end{enumerate}
\end{proof}

The gradually typed lambda calculus we consider is call-by-value
gradual lambda calculus with $\ra$ and $\nat$ as the only base types.
Casts are generated by the rule
\begin{mathpar}
  \inferrule
  {\Gamma \vdash M : A \and c : A \ltdyn A'}
  {\Gamma \vdash \upc c M : A'}

  \inferrule
  {\Gamma \vdash M : A' \and c : A \ltdyn A'}
  {\Gamma \vdash \dnc c M : A}
\end{mathpar}

Type precision is a binary relation on typed terms. The original
gradual guarantee rules are as follows:
\begin{mathpar}
  \inferrule
  {\Gamma^\ltdyn \vdash M \ltdyn M' : c \and
    c : A \ltdyn A'\and
    c' : A \ltdyn A_2'
  }
  {\Gamma^\ltdyn \vdash M \ltdyn (M :: A_2') : c'}

  \inferrule
  {\Gamma^\ltdyn \vdash M \ltdyn M' : c \and
    c : A \ltdyn A'\and
    c' : A_2 \ltdyn A'
  }
  {\Gamma^\ltdyn \vdash (M :: A_2) \ltdyn M' : c'}
\end{mathpar}
These two rules are admissible from the following principles:

%% \begin{mathpar}
%%   \inferrule
%%   {}
%%   {\dnc {\injarr{}} \upc {\injarr{}} M \equidyn M}

%%   %% \inferrule
%%   %% {}
%%   %% {}
%% \end{mathpar}

### 2024-01-23 paper-new/categorical-model.tex
sketchy stuff in the appendix on syntax and update to the extensional model

In particular, let $c : A \rel A'$ and $d : A' \rel A''$,
and let $f \in \mathcal V_f(A, A')$ and $g \in \mathcal V_f(A', A'')$.
We say that the pair $(c,d)$ is \emph{left-representable by} $(f, g)$
if the following two squares commute:

\begin{center}
  \begin{tabular}{ m{14em} m{14em} } 
    % UpL
    \begin{tikzcd}[ampersand replacement=\&]
      A \& {A'} \& {A''} \\
      {A'} \&\& {A''}
      \arrow["f"', from=1-1, to=2-1]
      %
      \arrow[from=1-3, to=2-3, Rightarrow, no head]
      %
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
      %
      \arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
    \end{tikzcd}
    &
    % UpR
    \begin{tikzcd}[ampersand replacement=\&]
      A \&\& {A'} \\
      {A} \& {A'} \& {A''}
      \arrow[from=1-1, to=2-1, Rightarrow, no head]
      %
      \arrow["g", from=1-3, to=2-3]
      %
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-3]
      %
      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \arrow["d"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
    \end{tikzcd}
  \end{tabular}
\end{center}

Dually, let $c \in \mathcal E_e(B, B')$ and $d \in \mathcal E_e(B', B'')$,
and let $f \in \mathcal E_f(B'', B')$ and $g \in \mathcal E_f(B', B)$.
We say the pair $(c,d)$ is \emph{right-representable by} $(f,g)$ if the
following two squares commute:

\begin{center}
  \begin{tabular}{ m{14em} m{14em} } 
    % DnR
    \begin{tikzcd}[ampersand replacement=\&]
      {B} \& {B'} \& {B''} \\
      {B} \&\& {B'}
      \arrow[from=1-1, to=2-1, Rightarrow, no head]
      %
      \arrow["f", from=1-3, to=2-3]
      %
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
      %
      \arrow["c"', from=2-1, to=2-3, no head]
    \end{tikzcd}
    &
    % DnL
    \begin{tikzcd}[ampersand replacement=\&]
      {B'} \&\& {B''} \\
      {B} \& {B'} \& {B''}
      \arrow["g"', from=1-1, to=2-1]
      %
      \arrow[from=1-3, to=2-3, Rightarrow, no head]
      %
      \arrow["d", from=1-1, to=1-3, no head]
      %
      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \arrow["d"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
    \end{tikzcd}
  \end{tabular}
\end{center}

Then we formulate the relationship between value relation morphisms and
function morphisms as follows:
\begin{enumerate}
  \item There is a functor $\upf : \mathcal V_e \to \mathcal V_f$ that is the
  identity on objects, such that $\upf_*(\mathcal V_e)$, the essential image of
  $\mathcal V_e$ under $\upf$, is thin.
  The objects of $\upf_*(\mathcal V_e)$ are the objects of $\mathcal V_f$, and
  the hom-set
  $\upf_*(\mathcal V_e)(A, A') = \{ f \in \mathcal V_f(A,A') \mid \exists c \in \mathcal V_e(A,A'). \upf(c) = f \}$.

  \item Every pair of morphisms $(c,d) \in \mathcal V_e(A, A') \times \mathcal V_e(A', A'')$ is
  left-representable by $(\upf(c), \upf(d))$.
\end{enumerate}

And likewise for computations:
\begin{enumerate}
  \item There is a functor $\dnf : \mathcal E_e^{op} \to \mathcal V_f$ that is
  the identity on objects, such that the essential image of $\mathcal E_e$ under
  $\dnf$ is thin.

  \item Every pair of morphisms $(c,d) \in \mathcal E_e(B, B') \times \mathcal E_e(B',B'')$
  is right-representable by $(\dnf(d), \dnf(c))$.
\end{enumerate}

\textbf{TODO: do we still need this?}
We also want something like
\[ F_c : \mathcal V_u^{op} \to \mathcal E_d \]
\[ U_c : \mathcal E_d^{op} \to \mathcal V_u \]
which ensures that if $R$ is a value edge equivalent to $A(u,-)$ then
\[ F(R) = F(A(u,-)) = (F A)(-,F u) \]

In summary, an extensional model consists of:
\begin{enumerate}
  \item CBPV models $\mathcal M_f$ and $\mathcal M_{sq}$
  \item CBPV morphisms $r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$
  \item Thinness: there is at most one square with a given boundary
  \item A ``horizontal composition" operation on value relations and on computation relations
  (from which we can define the categories $\ve$ and $\ee$ of value types/relations and computation types/relations, repsectively).
  \item The categories $\ve$ and $\ee$ are thin up to an identity square
  \item A functor $\upf : \mathcal V_e \to \mathcal V_f$ that is the identity on objects,
  such that $\upf_*(\mathcal V_e)$, the essential image of $\mathcal V_e$ under $\upf$, is thin.
  \item Every pair of morphisms $(c,d) \in \mathcal V_e(A, A') \times \mathcal V_e(A', A'')$ is
  left-representable by $(\upf(c), \upf(d))$.
  \item A functor $\dnf : \mathcal E_e^{op} \to \mathcal V_f$ that is
  the identity on objects, such that the essential image of $\mathcal E_e$ under $\dnf$ is thin.
  \item Every pair of morphisms $(c,d) \in \mathcal E_e(B, B') \times \mathcal E_e(B',B'')$
  is right-representable by $(\dnf(d), \dnf(c))$.
\end{enumerate}

>>>>

In particular, let $c : A \rel A'$ and $f \in \mathcal V_f(A, A')$. We
say that $c$ is \emph{left-representable by} $f$ if for any $c_l : A_l
\rel A$ and $c_r : A' \rel A_r$ the following squares commute\max{is ``commute'' good terminology?}:
\begin{center}
  \begin{tabular}{ m{14em} m{14em} } 
    % UpL
    \begin{tikzcd}[ampersand replacement=\&]
      A \& {A'} \& {A_r} \\
      {A'} \&\& {A_r}
      \arrow["f"', from=1-1, to=2-1]
      %
      \arrow[from=1-3, to=2-3, Rightarrow, no head]
      %
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["c_r", "\shortmid"{marking}, no head, from=1-2, to=1-3]
      %
      \arrow["c_r"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
    \end{tikzcd}
    &
    % UpR
    \begin{tikzcd}[ampersand replacement=\&]
      A_l \&\& {A} \\
      {A_l} \& {A} \& {A'}
      \arrow[from=1-1, to=2-1, Rightarrow, no head]
      %
      \arrow["f", from=1-3, to=2-3]
      %
      \arrow["c_l", "\shortmid"{marking}, no head, from=1-1, to=1-3]
      %
      \arrow["c_l"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \arrow["c"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
    \end{tikzcd}
  \end{tabular}
\end{center}

Dually, let $d \in B \rel B'$ and $\phi \in \mathcal E_f(B', B)$.  We
say that $d$ is \emph{right-representable by} $\phi$ if for any $d_l :
B_l \rel B$ and $d_r : B' \rel B_r$ the following two squares commute:

\begin{center}
  \begin{tabular}{ m{14em} m{14em} } 
    % DnR
    \begin{tikzcd}[ampersand replacement=\&]
      {B_l} \& {B} \& {B'} \\
      {B_l} \&\& {B}
      \arrow[from=1-1, to=2-1, Rightarrow, no head]
      %
      \arrow["\phi", from=1-3, to=2-3]
      %
      \arrow["d_l", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
      %
      \arrow["d_l"', from=2-1, to=2-3, no head]
    \end{tikzcd}
    &
    % DnL
    \begin{tikzcd}[ampersand replacement=\&]
      {B'} \&\& {B_r} \\
      {B} \& {B'} \& {B_r}
      \arrow["\phi"', from=1-1, to=2-1]
      %
      \arrow[from=1-3, to=2-3, Rightarrow, no head]
      %
      \arrow["d_r", from=1-1, to=1-3, no head]
      %
      \arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \arrow["d_r"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
    \end{tikzcd}
  \end{tabular}
\end{center}

Then we formulate the relationship between relation morphisms and
function morphisms as follows:
\begin{enumerate}
\item There is an identity-on-objects functor $\upf : \mathcal V_e \to
  \mathcal V_f$ such that every $c$ is left-representable by $\upf(c)$.
\item There is an identity-on-objects functor $\dnf : \mathcal
  E_e^{op} \to \mathcal \mathcal E_f$ such that every $d$ is
  right-representable by $\dnf(d)$.
\end{enumerate}

%% \textbf{TODO: do we still need this?}
%% We also want something like
%% \[ F_c : \mathcal V_u^{op} \to \mathcal E_d \]
%% \[ U_c : \mathcal E_d^{op} \to \mathcal V_u \]
%% which ensures that if $R$ is a value edge equivalent to $A(u,-)$ then
%% \[ F(R) = F(A(u,-)) = (F A)(-,F u) \]

In summary, an extensional model consists of:
\begin{enumerate}
  \item A CBPV model internal to reflexive graphs.
  \item Composition and identity on value and computation relations that form a category.
  \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that left-represents it.
  \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that right-represents it.
  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations
    \begin{itemize}
    \item $U(dd') = U(d)U(d')$
    \item $F(cc'') = F(c)F(c')$
    \item $(cc') \to (dd') = (c \to d)(c' \to d')$
    \item $(c_1c_1') \times (c_2c_2') = (c_1 \times c_2)(c_1'\times c_2')$
    \end{itemize}
  \item A distinguished value type $D$ with distinguished relation
    $\injarr{}: U(D \to F D) \rel D$ satisfying $\dnc {\injarr{}}F(\upc{\injarr{}}) = \id$
\end{enumerate}
